Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of neutron star - black hole binary observations

### 2016-10-11 Paper/paper.tex
first bout of changes from Bens comments

% \red{
% BEgin this paragraph by pointing out that EOS is universal among all NSs. 
% Therefore, one universal $\lambdans$ vs $\mns$ curve. Therefore, by combining
% multiple BHNS events, expect to measure $\lambdans$ better than with any 
% single BHNS event.
% }
% % \prayush{
The EoS of matter at nuclear densities is believed to be universal among all
neutron stars. The Tolman-Oppenheimer-Volkoff equation~\cite{Tolman:1939jz,
Oppenheimer:1939ne,1934PNAS...20..169T}
would then predict that NS properties satisfy a universal relationship
between $\lambdans$ and $\mns$. As the final part of
this paper, we combine information from multiple observations of realistic NSBH
systems and perform a fully-Bayesian analysis of how our estimation of
$\lambdans$ changes as we accumulate detections. This is similar to an earlier
study~\cite{DelPozzo:13} aimed at binary neutron stars.
% % }
% % 
% \red{
% MOVE into later sections. Keep here only 1-2 informational sentences.
% }
% % 
We restrict ourselves to a population of NSs with masses clustered very tightly
around $1.35M_\odot$ (with a negligible variance), and negligible spins. We
sample different nuclear EoSs by sampling entire populations fixing different
values for the NS tidal deformability.
% These values correspond to a hard EoS at the lower end, and a soft EoS at the upper.
For all populations, we take source locations to be uniformly distributed in
spatial volume, and source orientations to be uniform on the $2-$sphere. To
summarize, we find the following:
% 
(a) Our best estimate for $\lambdans$ improves much faster than our statistical
uncertainties. Within $10-20$ detections, the median value of
$\lambdans$ lies within $10\%$ of the actual value.
(b) Measurement uncertainties for $\lambdans$, on the other hand, depend on
$\lambdans$ itself. We find that for hard equations of state (with 
$\lambdans\geq 1000$), $10-20$ observations are sufficient to constrain
$\lambdans$ within $\pm 50\%$. For softer equations of state, the same level
of certainty would require substantially more ($25-40$) observations.
% 
(c) Further, if the astrophysical ``mass-gap''~\cite{Bailyn:1997xt,
Kalogera:1996ci,Kreidberg:2012,Littenberg:2015tpa} is real, we find that $20-50\%$
additional observations would be required to attain the same measurement
accuracy as above. And, (d) putting tighter constraints on the $\lambdans$ of a
population would require $50+$ NSBH observations, in any scenario.
% 
All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Techniques}\label{s1:techniques}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% #################
% THESE FIGURES ARE FOR THE NEXT SECTION
\begin{figure*}
\centering 
\textbf{Illustrative posterior probability distributions for mass-ratio $\eta$,}\par
\textbf{at different SNR values (shown in legend).}\par\medskip
\includegraphics[width=1.8\columnwidth]{plots/SingleSystemEta_q4_0_mc2_25_chi0_50}
\caption{We show here probability distributions for mass ratio $\eta$ as measured
for the same signal at different SNRs. The intrinsic parameters of the source
are: $q = \mbh/\mns = 5.4M_\odot/1.35M_\odot = 4$, $\chibh=+0.5$, and $\lambdans=2000$;
and the signal is injected at SNRs $\rho=\{20,30,50\}$ (left to right). The templates
{\it ignore} tidal effects.
% 
In each panel: the dashed red line marks the median value
$\eta^\mathrm{Median}$, while the dashed green line show the true value
$\eta^\mathrm{Injected}$. The darker shading shows
the recovered $90\%$ credible interval for $\eta$, $(\Delta\eta)^{90\%}$.
% 
Comparing systematic and statistical errors, we find that:
at $\rho=20$, $\eta$ measurement is dominated by statistical
errors; at $\rho=30$, the two become comparable; and 
for louder signals ($\rho\simeq50$), the systematic errors dominate.
}
\label{fig:SingleSystemEtaPDFvsSNR}
\end{figure*}
% 
% ###############################
% In this section we summarize various technical aspects of this paper. The model
% used to obtain tidal waveforms is described in Sec.~\ref{s2:waveforms}. The 
% process of inferring source parameters from a GW signal is summarized in 
% Sec.~\ref{s2:bayesian}. The process of generating and combining multiple events
% is instead detailed in the corresponding results section,
% Sec.~\ref{s2:astro_multiple}.

@@

@@

\textbf{Paradigm A \hspace{6cm} Paradigm B}\par\medskip

@@

In Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0} we show illustrative results
for a single population with neutron star deformability $\lambdans=800$.
In the left panel, each curve shows the 
probability distributions for $\lambdans$ as inferred from $N$ events, with $N$
ranging from $1-80$. We also mark the $90\%$ credible intervals associated
with each of the probability distribution curves. The first few observations
do not have
enough information to bound $\lambdans$ much more than
our prior from Eq.~\ref{eq:lprior} does. However, the median of these
cumulative distributions appears to track the true value of $\lambdans$ 
rapidly.
% 
In the right panel, we present information derived from the left panel.
The line-circle curve shows the measured median value from $N$ observations.
The pair of dashed (dotted) horizontal lines mark
$\pm25\%$ ($\pm50\%$) error bars. At each $N$, the range spanned by the 
filled region is the $90\%$ credible interval deduced from the same 
events. This figure somewhat quantifies the qualitative deductions we made
from the left panel. We find that the median does track the true value
quickly, reaching $10\%$ accuracy with $10-15$ observations. With the same
information, our credible intervals also shrink to $\pm 25\%$.
% 
In Fig.~\ref{fig:TT_Lambda_vs_N_CI90_0} we show further results from four
independent populations for $\lambdans=\{500,1000,1500,2000\}$. As in the 
right panel of Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}, the line-circle curves
track the median $\lambdans$, while the filled regions show
the associated $90\%$ credible intervals. From the figure, we observe
the following: (i) the shrinkage of credible interval widths with increasing
$N$ happens in a similar manner for each $\lambdans$, (ii) within $\sim 10$
observations, median $\lambdans$ for all populations lie within
$10\%$ of the true values, and (iii) it takes
approximately $20$ events to distinguish definitively (with $90\%$ credibility)
between deformable NSs with $\lambdans=2000$ and compact NSs with 
$\lambdans=500$, or equivalently to distinguish between hard, moderate and soft
nuclear equations of state. This is comparable to what has been found for
binary neutron stars~\cite{DelPozzo:13,Chatziioannou:2015uea,Agathos:2015a}.

So far we have discussed individual realizations of NSBH populations. The 
underlying stochasticity of the population generation process makes it
difficult to draw generalized inferences (from a single realization of an
NSBH population) about the measurability of $\lambdans$. In order to mitigate
this, we discuss ensembles of population draws next. In
Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} we show the median
$\lambdans$ as a function of the number of observed
events, for four population ensembles, with a hundred population draws
in each ensemble. Lets focus on the {\it top left} panel first. In it, we
show the median $\lambdans$ for all populations in four ensembles,
with true $\lambdans=\{2000,1500,1000,500\}$ from top to bottom.
Populations highlighted in color are simply those that we discussed in
Fig.~\ref{fig:TT_Lambda_vs_N_CI90_0}. Dash-dotted horizontal lines
demarcate $\pm10\%$ error intervals around the true $\lambdans$ values.
The panel just below it shows the range of $\lambdans$ that encloses the
median $\lambdans$ for $90\%$ of the populations in {\it each}
ensemble. In other words, this panel shows the range of $\lambdans$ within which
the median $\lambdans$ value for $90\%$ of NSBH populations is
expected to lie. From these panels, we conclude that our median $\lambdans$
values will be within $10\%$ of the {\it true} value after $\sim 25$
detections of less deformable neutron stars ($\lambdans\leq 1000$), or
after as few as $15$ detections of more deformable neutron stars
($\lambdans\geq 1500$).
% 
The results discussed in Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0},
\ref{fig:TT_Lambda_vs_N_CI90_0} and the left two panels
of Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} apply to the parameter
distribution spanned by the grid G. This distribution allows for $\mbh$
as low as $2.7M_\odot$ (i.e. $q=2$).
Given that disruptive signatures are strongest for small $\mbh$, we now
investigate an alternate paradigm in which no black hole masses fall within
the mass gap $2-5M_\odot$ suggested by astronomical
observations~\cite{Bailyn:1997xt,Kalogera:1996ci,Kreidberg:2012,
Littenberg:2015tpa}. We will henceforth denote our standard paradigm, which
does not respect the mass-gap, as paradigm A; with paradigm B being
this alternate scenario.
% 
Both right panels of the figure are identical to their corresponding left
panels, but drawn under population paradigm B. Under this paradigm, we
expectedly find that information accumulation is much slower. It would
take $25-40$ detections with $\rho\geq10$ under this paradigm, for our median
$\lambdans$ to converge within $10\%$ of its true value.
% % 
% This is a promising deduction, as one might expect to see $\mathcal{O}(10)$
% events over a few years' timescale with design aLIGO~\cite{Abadie:2010cfa}.

@@

We probe two astrophysical paradigms, one that allows for BH masses to lie
within the astrophysical mass-gap (paradigm A), and one that does not (paradigm
B).
% 
{\it For paradigm A}, we find the following: (i) for the softer equations of state
that result in less deformable neutron stars, $15-20$ detections are sufficient
for our median value of $\lambdans$ to lie within $10\%$ of the true
value. (ii) For NSBH populations with more deformable NSs ($\lambdans> 1000$),
the same is achievable within as few as $10$ (or $15$ at most) realistic
observations. (iii) The statistical uncertainty associated with $\lambdans$
measurement can be restricted to be within $\pm50\%$ using $10-20$ observations
when $\lambdans> 1000$), and using $25-40$ observations for softer equations
of state. All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}, if astrophysical BHs are allowed
masses $< 5M_\odot$ (i.e. in the mass-gap). However, further
restricting $\lambdans$ will require $50+$ NSBH observations.
% 
{\it For paradigm B}, we find the information accumulation to be somewhat slower.
While the quantitative inferences for populations with $\lambdans>1000$ are
not affected significantly, we find that $\lambdans< 1000$ populations require 
$10-20\%$ more events to attain the same measurement accuracy as under
paradigm A. In either case, the accumulation of information from NSBH signals
is likely slower than from BNS inspirals~\cite{Mandel:2009pc,Lackey2014,
Wade:2014vqa}, although template models for the latter may be more
uncertain due to missing point-particle PN terms at orders comparable to
the tidal terms.
% 
We conclude that within as few as $20-30$ observations of disruptive NSBH
mergers, aLIGO will begin to place interesting bounds on NS deformability.
This, amongst other things, will allow us to rank different equations of 
state for neutron star matter from most to least likely, within a few years'
detector operation.
% 
Our methods and results are detailed in Sec.~\ref{s1:multiple_observations}.

>>>>

% \red{
% BEgin this paragraph by pointing out that EOS is universal among all NSs. 
% Therefore, one universal $\lambdans$ vs $\mns$ curve. Therefore, by combining
% multiple BHNS events, expect to measure $\lambdans$ better than with any 
% single BHNS event.
% }
% % \prayush{
The EoS of matter at nuclear densities is believed to be universal among all
neutron stars. The Tolman-Oppenheimer-Volkoff equation~\cite{Tolman:1939jz,
Oppenheimer:1939ne,1934PNAS...20..169T}
would then predict that NS properties satisfy a universal relationship
between $\lambdans$ and $\mns$. As the final part of
this paper, we combine information from multiple observations of realistic NSBH
systems and perform a fully-Bayesian analysis of how our estimation of
$\lambdans$ changes as we accumulate detections. This is similar to an earlier
study~\cite{DelPozzo:13} aimed at binary neutron stars.
% % }
% % 
% \red{
% MOVE into later sections. Keep here only 1-2 informational sentences.
% }
% % 
We restrict ourselves to a population of NSs with masses clustered very tightly
around $1.35M_\odot$ (with a negligible variance), and negligible spins. We
sample different nuclear EoSs by sampling entire populations fixing different
values for the NS tidal deformability.
% These values correspond to a hard EoS at the lower end, and a soft EoS at the upper.
For all populations, we take source locations to be uniformly distributed in
spatial volume, and source orientations to be uniform on the $2-$sphere. To
summarize, we find the following:
% 
(a) Our median estimate for $\lambdans$ starts out prior dominated, but 
converges to within $10\%$ of the true value within $10-20$ detections.
(b) Measurement uncertainties for $\lambdans$, on the other hand, depend on
$\lambdans$ itself. We find that for hard equations of state (with 
$\lambdans\geq 1000$), $10-20$ observations are sufficient to constrain
$\lambdans$ within $\pm 50\%$. For softer equations of state, the same level
of certainty would require substantially more ($25-40$) observations.
% 
(c) Further, if the astrophysical ``mass-gap''~\cite{Bailyn:1997xt,
Kalogera:1996ci,Kreidberg:2012,Littenberg:2015tpa} is real, we find that $20-50\%$
additional observations would be required to attain the same measurement
accuracy as above. And, (d) putting tighter constraints on the $\lambdans$ of a
population would require $50+$ NSBH observations, in any scenario.
% 
All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Techniques}\label{s1:techniques}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% #################
% THESE FIGURES ARE FOR THE NEXT SECTION
\begin{figure*}
\centering 
\textbf{Illustrative posterior probability distributions for mass-ratio $\eta$,}\par
\textbf{at different SNR values (shown in legend).}\par\medskip
\includegraphics[width=1.8\columnwidth]{plots/SingleSystemEta_q4_0_mc2_25_chi0_50}\\
\caption{We show here probability distributions for mass ratio $\eta$ as measured
for the same signal at different SNRs. The intrinsic parameters of the source
are: $q = \mbh/\mns = 5.4M_\odot/1.35M_\odot = 4$, $\chibh=+0.5$, and $\lambdans=2000$;
and the signal is injected at SNRs $\rho=\{20,30,50\}$ (left to right). The templates
{\it ignore} tidal effects.
% 
In each panel: the dashed red line marks the median value
$\eta^\mathrm{Median}$, while the dashed green line show the true value
$\eta^\mathrm{Injected}$. The darker shading shows
the recovered $90\%$ credible interval for $\eta$, $(\Delta\eta)^{90\%}$.
% 
Comparing systematic and statistical errors, we find that:
at $\rho=20$, $\eta$ measurement is dominated by statistical
errors; at $\rho=30$, the two become comparable; and 
for louder signals ($\rho\simeq50$), the systematic errors dominate.
}
\label{fig:SingleSystemEtaPDFvsSNR}
\end{figure*}
% 
% ###############################
% In this section we summarize various technical aspects of this paper. The model
% used to obtain tidal waveforms is described in Sec.~\ref{s2:waveforms}. The 
% process of inferring source parameters from a GW signal is summarized in 
% Sec.~\ref{s2:bayesian}. The process of generating and combining multiple events
% is instead detailed in the corresponding results section,
% Sec.~\ref{s2:astro_multiple}.

@@

\includegraphics[width=0.8\columnwidth]{plots/TNMchirpBiasesOverCIWidthsVsSNR_Lambda500_CI90_0}
% \includegraphics[width=0.67\columnwidth]{plots/TNMchirpBiasesOverCIWidthsVsSNR_Lambda800_CI90_0}
\includegraphics[width=0.8\columnwidth]{plots/TNMchirpBiasesOverCIWidthsVsSNR_Lambda1000_CI90_0}\\
\includegraphics[width=0.8\columnwidth]{plots/TNMchirpBiasesOverCIWidthsVsSNR_Lambda1500_CI90_0}
\includegraphics[width=0.8\columnwidth]{plots/TNMchirpBiasesOverCIWidthsVsSNR_Lambda2000_CI90_0}

@@

\textbf{No Mass-Gap \hspace{6cm} Mass-Gap}\par\medskip

@@

In Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0} we show illustrative results
for a single population with neutron star deformability $\lambdans=800$.
In the top panel, each curve shows the 
probability distributions for $\lambdans$ as inferred from $N$ events, with $N$
ranging from $1-80$. We also mark the $90\%$ credible intervals associated
with each of the probability distribution curves. The first few observations
do not have enough information to bound $\lambdans$ much more than
our prior from Eq.~\ref{eq:lprior} does. 
% 
In the bottom panel, we present information derived from the top panel.
The line-circle curve shows the measured median value from $N$ observations.
The pair of dashed (dotted) horizontal lines mark
$\pm25\%$ ($\pm50\%$) error bars. At each $N$, the range spanned by the 
filled region is the $90\%$ credible interval deduced from the same 
events. This figure somewhat quantifies the qualitative deductions we made
from the left panel. We find that the median does track the true value quickly,
reaching within its $10\%$ with $10-15$ observations. This is as one expects of 
injections in zero noise where random fluctuations are unable to shift the
median away from the true value, so long as the measurement is not restricted
by the prior. With the same information, our credible intervals also shrink to $\pm 25\%$.
% 
In Fig.~\ref{fig:TT_Lambda_vs_N_CI90_0} we show further results from four
independent populations for $\lambdans=\{500,1000,1500,2000\}$. As in the 
right panel of Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}, the line-circle curves
track the median $\lambdans$, while the filled regions show
the associated $90\%$ credible intervals. From the figure, we observe
the following: (i) the shrinkage of credible interval widths with increasing
$N$ happens in a similar manner for each $\lambdans$, 
% (ii) within $\sim 10$ observations, median $\lambdans$ for all populations lie
% within $10\%$ of the true values, 
and (ii) it takes
approximately $20$ events to distinguish definitively (with $90\%$ credibility)
between deformable NSs with $\lambdans=2000$ and compact NSs with 
$\lambdans=500$, or equivalently to distinguish between hard, moderate and soft
nuclear equations of state. This is comparable to what has been found for
binary neutron stars~\cite{DelPozzo:13,Chatziioannou:2015uea,Agathos:2015a}.

So far we have discussed individual realizations of NSBH populations. The 
underlying stochasticity of the population generation process makes it
difficult to draw generalized inferences (from a single realization of an
NSBH population) about the measurability of $\lambdans$. In order to mitigate
this, we discuss ensembles of population draws next. In
Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} we show the median
$\lambdans$ as a function of the number of observed
events, for four population ensembles, with a hundred population draws
in each ensemble. Lets focus on the {\it top left} panel first. In it, we
show the median $\lambdans$ for all populations in four ensembles,
with true $\lambdans=\{2000,1500,1000,500\}$ from top to bottom.
Populations highlighted in color are simply those that we discussed in
Fig.~\ref{fig:TT_Lambda_vs_N_CI90_0}. Dash-dotted horizontal lines
demarcate $\pm10\%$ error intervals around the true $\lambdans$ values.
The panel just below it shows the range of $\lambdans$ that encloses the
median $\lambdans$ for $90\%$ of the populations in {\it each}
ensemble. In other words, this panel shows the range of $\lambdans$ within which
the median $\lambdans$ value for $90\%$ of NSBH populations is
expected to lie. From these panels, we observe that our median $\lambdans$
values will be within $10\%$ of the {\it true} value after $\sim 25$
detections of less deformable neutron stars ($\lambdans\leq 1000$), or
after as few as $15$ detections of more deformable neutron stars
($\lambdans\geq 1500$). This is not surprising because we inject simulated
signals in {\it zero} noise, which ensures that the median not be shifted away
from the true value. That it takes $15+$ events for the median to approach
the true value is a manifestation of the fact that the measurement is limited
by the prior on $\lambdans$ when we have fewer than $15$ events.
% 
The results discussed in Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0},
\ref{fig:TT_Lambda_vs_N_CI90_0} and the left two panels
of Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} apply to the parameter
distribution spanned by the grid G. This distribution allows for $\mbh$
as low as $2.7M_\odot$ (i.e. $q=2$).
Given that disruptive signatures are strongest for small $\mbh$, we now
investigate an alternate paradigm in which no black hole masses fall within
the mass gap $2-5M_\odot$ suggested by astronomical
observations~\cite{Bailyn:1997xt,Kalogera:1996ci,Kreidberg:2012,
Littenberg:2015tpa}. We will henceforth denote our standard paradigm, which
does not respect the mass-gap, as paradigm A; with paradigm B being
this alternate scenario.
% 
Both right panels of the figure are identical to their corresponding left
panels, but drawn under population paradigm B. Under this paradigm, we
expectedly find that information accumulation is much slower. It would
take $25-40$ detections with $\rho\geq10$ under this paradigm, for our median
$\lambdans$ to converge within $10\%$ of its true value.
% % 
% This is a promising deduction, as one might expect to see $\mathcal{O}(10)$
% events over a few years' timescale with design aLIGO~\cite{Abadie:2010cfa}.

@@

We probe two astrophysical paradigms, one that allows for BH masses to lie
within the astrophysical mass-gap (paradigm A), and one that does not (paradigm
B).
% 
{\it For paradigm A}, we find the following: (i) for the softer equations of
state that result in less deformable neutron stars, $15-20$ detections bring
the measured probability distribution for $\lambdans$ entirely within the prior,
which ensures that the median $\lambdans$ tracks the true value to within $10\%$.
(ii) For NSBH populations with more deformable NSs ($\lambdans> 1000$),
the same is achievable within as few as $10$ (or $15$ at most) realistic
observations. (iii) The statistical uncertainty associated with $\lambdans$
measurement can be restricted to be within $\pm50\%$ using $10-20$ observations
when $\lambdans> 1000$), and using $25-40$ observations for softer equations
of state. All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}, if astrophysical BHs are allowed
masses $< 5M_\odot$ (i.e. in the mass-gap). However, further
restricting $\lambdans$ will require $50+$ NSBH observations.
% 
{\it For paradigm B}, we find the information accumulation to be somewhat slower.
While the quantitative inferences for populations with $\lambdans>1000$ are
not affected significantly, we find that $\lambdans< 1000$ populations require 
$10-20\%$ more events to attain the same measurement accuracy as under
paradigm A. In either case, the accumulation of information from NSBH signals
is likely slower than from BNS inspirals~\cite{Mandel:2009pc,Lackey2014,
Wade:2014vqa}, although template models for the latter may be more
uncertain due to missing point-particle PN terms at orders comparable to
the tidal terms.
% 
We conclude that within as few as $20-30$ observations of disruptive NSBH
mergers, aLIGO will begin to place interesting bounds on NS deformability.
This, amongst other things, will allow us to rank different equations of 
state for neutron star matter from most to least likely, within a few years'
detector operation.
% 
Our methods and results are detailed in Sec.~\ref{s1:multiple_observations}.
